---
title: "Objects and functions"
format: html
editor: visual
---

## Functions applied to objects

In R, data objects are structures that store data that can then be acted on by a function. This is best understood with a simple example. Consider the a set of numbers that have been assigned to a variables called, "my_numbers".

```{r}
my_numbers <- c(3, 5, 8, 2)
```

We've created a vector (one of many types of data objects) that can be easily called by making reference to its designated name.

```{r}
my_numbers
```

If we wanted to calculate the sum of those numbers, we could simple apply the **`sum`** function to that object as follows:

```{r}
sum(my_numbers)
```

In R, there are several data objects that you will commonly use. These include:

1.  **`Vectors`**: Vectors are the most basic data object in R. They can contain elements of the same types, such as numeric, character, and logical.
2.  **`Data frames`**: Data frames are tables of data where each column can contain elements of different types.
3.  **`Table`**:
4.  **`Lists`**: Lists are similar to vectors, but they can contain elements of different types and can also contain other lists.
5.  **`Matrices`**: Matrices are two-dimensional arrays of data. All elements in a matrix must be of the same type.
