---
title: "Welcome to the tidyverse: installing and using packages"
editor: visual
---

```{r, echo=FALSE}
library(tidyverse)
```

So, in R programming, a package is basically a bunch of extra functions, data, and documentation that can be used to extend the capabilities of base R. You can think of it as extending the vocabulary of R (giving you more functions to work with). Packages may also contain datasets that you can use to practice with.

To use a package in R, you **first** have to install it. You can do this using the **`install.packages()`** function and passing in the name of the package as the argument. For example, if you want to install the **`dplyr`** package, you would run:

```{r}
install.packages("dplyr")
```

Once you've installed a package, you can load it into your current R session using the **`library()`** function. This will make all the functions, data, and documentation in the package available for you to use. So, if you want to load the **`dplyr`** package, you would run:

```{r}
library(dplyr)
```

You can also install and load a package in one step by using the **`require()`** function. This will install the package if it's not already installed, and then load it into your current session. This is a useful function if you're writing code that will be run on someone elses machine on which the packages might not have been installed. For example:

```{r}
require(dplyr)
```

Once you've installed a package you can access any data that is included in that package and utilize the functions that come with the package. So in the case of the 'dplyr' page you would run:

```{r}
select(mpg, manufacturer, year, class)
```

If there is any chance that the same text for a function in more than one package (leading to a potential conflict) then its a good idea to include the package name when calling the function like this:

```{r}
dplyr::filter(mpg, manufacturer, year, class)
```

### The tidyverse packages

You are going to love the tidyverse. The tidyverse is a group of R packages that work together to make it easier to wrangle and visualize data. All the packages in the tidyverse have a similar syntax and are designed to be easy to use.

Some of the main packages in the tidyverse include:

-   **`ggplot2`**: for creating visualizations

-   **`dplyr`**: for data manipulation

-   **`tidyr`**: for cleaning and reshaping data

-   **`readr`**: for reading and writing data

-   **`purrr`**: for functional programming

-   **`tibble`**: for working with tibbles (a modified version of data frames)

-   **`stringr`**: for working with strings

-   **`lubridate`**: for working with dates and times

There are other packages that are part of the tidyverse, but these are the core ones.

To get the tidyverse package, simply run:

```{r}
install.packages("tidyverse")
library(tidyverse)
```

The tidyverse is really useful for data manipulation and visualization in R and is used by a most R users today. We highly recommend working with these packages and will be using them for the remainder of this book.
